### 2013-02-07 doc/xps2013/proposal.tex
done for the day

Irregular applications generate tasks with work, interdependences, or memory accesses that are highly sensitive to input. Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow. Important contemporary examples include processing large graphs in the business, national security, machine learning, data-driven science, and social network computing domains. For these emerging applications, fast response -- given the sheer amount of data -- requires multinode systems. The most broadly available multinode systems are those built from x86 commodity computing nodes interconnected via ethernet or infiniband. Our goal is to enable scalable performance of irregular applications on these mass market systems.  Our goal is challenging for two key reasons:

\noindent{\bf Irregular applications exhibit little spatial locality.} It is not atypical for any given task's data references to be spread randomly across the entire memory of the system. This makes current memory hierarchy features ineffective. Caches are of little assistance with such low data re-use and spatial locality. Commodity prefetching hardware is effective only when addresses are known many cycles before the data is consumed or the accesses follow a predictable pattern, neither of which occurs in irregular applications. As a consequence, commodity microprocessors stall often when executing irregular applications.

\noindent{\bf Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node. Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks. Injection rate into the network is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.

@@

Luckily, many of the important irregular applications (e.g., graph processing, machine learning, ...) naturally offer large amounts of concurrency. This immediately suggests taking advantage of concurrency to tolerate the latency of data movement. The fully custom Tera MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular applications by using concurrency to hide latencies. It had a large distributed shared memory with no caches.  On every clock cycle, each processor would execute a ready instruction chosen from one of its 128 hardware thread contexts, a sufficient number to fully hide memory access latency.  The network was designed with a single word injection rate that matched the processor clock frequency and sufficient bandwidth to sustain a reference from every processor on every clock cycle. Unfortunately, while an excellent match to extremely irregular applications, the MTA was not cost-effective on applications that could exploit locality and had very poor single-thread performance, making it a commercial failure. The Cray XMT approximates the Tera MTA-2, substantially reducing its cost but not overcoming its narrow range of applicability.

@@

Our preliminary work on Grappa shows that it runs several graph-crunching applications (classic examples of irregular behavior) very efficiently on a commodity cluster.  For unbalanced tree search, Grappa is over 3X faster and shows greatly improved scalability compared to the Cray XMT; conversely, for breadth first search Grappa is 2.5X slower.  Most importantly, however, for significantly less real world cost, users can \emph{add} significantly more processors to a commodity cluster than an XMT machine and use Grappa to achieve scalable performance.

@@

\begin{itemize}
\item Presently, applications are hand-coded to the Grappa interface.  A significant research thrust of the proposed work is to ease this programming burden by language and compilation support.

\item Our preliminary work has also revealed limitations of the current Grappa runtime system with regards to network scaling.  This proposal seeks funding to enhance the network layer and explore the hardware/software interactions of remote direct memory access (RDMA).

\item The size of data being processed by large systems today (hundreds of terabytes) dwarfs what can fit into a single system. Multiple system nodes are required to maintain and access all of this data efficiently.  However, even with hundreds of channels to persistent storage, loading this data into DRAM can prove a bottleneck.  The Grappa runtime is designed to make applications tolerate extremely high memory latency, thus an idea we wish to explore is using solid state disks as main memory.

\item Presently there does not exist a comprehensive benchmark suite of irregular and graph applications.  We also seeking funding through this proposal to put together just such a benchmark suite.  This suite will include benchmarks, reference implementations, and input data-sets.  In addition a website will be created \texttt{graphbench.org} to distribute the benchmark suite and publish results.

@@

@@

Delegate operations are always executed at the home core of their address, and while arbitrary memory operations can be delegated, we restrict the use of delegate operations in three ways to make them more useful for synchronization. First, we limit each task to one outstanding delegate operation to avoid the possibility of reordering in the network. Second, we limit delegate operations to operate on objects in the 2D address space or objects that fit in a single block of the linear address space so they can be satisfied with a single network request. Finally, no context switches are allowed while the data is being modified. Given these restrictions, we can ensure that delegate operations for the same address from multiple requesters are always serialized through a single core in the system, providing atomic semantics without using atomic operations. Figure~\ref{fig:delegate-cache} depicts an example of how delegate and cache operations interact.

\begin{figure*}[htb]
\begin{center}
  \includegraphics[width=6in]{figs/delegate-cache}
\begin{minipage}{\columnwidth}

\caption{\label{fig:delegate-cache} \textbf{Delegation and cache example:} In step 1, a core allocates an array in the global heap. It then spawns two tasks on remote cores to increment elements of the array.  The first task increments two elements of the array using cache operations. In step 2, the task is invoked. A cache request is issued for two adjacent integers starting at the second element of the array. Since these element are stored in the memories of two different cores, this requires the sending of two messages in step 3. The task is suspended until both responses arrive in step 4. The data carried in these responses is stored in the local buffer. The elements are then increment in the buffer. Then in step 5, the modified data is sent back to the home node. Acknoledgements are returned in step 6 so the task knows when the writes are complete. The second task increments an element of the array with a delegate operation. In step 7, the task is invoked. A delegate request is sent to the home core of the array element with the increment value. The task suspends until the response is received. In step 8, the increment is executed on the remote core. A response is returned in step 9 with the previous value of the array element. }

\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure*}

@@

The final challenge is that atomic operations via RDMA do not interact atomically with processors in a system.  Semantically it is not entirely clear what happens if a processor issues an atomic operation on a memory address at the same time the network card is attempting to service a remote atomic access. \comment{maybe this is not true.  have to think some more about PCIe an what happens here.  might not be a problem.  -M}

@@

\comment{mostly a copy and paste of the pomace?}

@@

@@

>>>>

Irregular applications generate tasks with work, interdependences, and memory accesses that are highly sensitive to input. Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow. Important contemporary examples include processing large graphs in the business, national security, machine learning, data-driven science, and social network computing domains. For these emerging applications, fast response -- given the sheer amount of data -- requires multinode systems. The most broadly available multinode systems are those built from x86 commodity computing nodes interconnected via ethernet or infiniband. Our goal is to enable scalable performance of irregular applications on these mass market systems.  Our goal is challenging for two key reasons:

\paragraph{Irregular applications exhibit little spatial locality.} It is not atypical for any given task's data references to be spread randomly across the entire memory of the system. This makes current memory hierarchy features ineffective. Caches are of little assistance with such low data re-use and spatial locality. Commodity prefetching hardware is effective only when addresses are known many cycles before the data is consumed or the accesses follow a predictable pattern, neither of which occurs in irregular applications. As a consequence, commodity microprocessors stall often when executing irregular applications.

\paragraph{Irregular applications frequently request small amounts of off-node data.} On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node. Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks. Injection rate into the network is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.

@@

Luckily, many of the important irregular applications naturally offer large amounts of concurrency. This immediately suggests taking advantage of concurrency to tolerate the latency of data movement. The fully custom Tera MTA-2~\cite{tera:mta1} system is a classic example of supporting irregular applications by using concurrency to hide latencies.  It had a large distributed shared memory with no caches.  On every clock cycle, each processor would execute a ready instruction chosen from one of its 128 hardware thread contexts, a sufficient number to fully hide memory access latency.  The network was designed with a single word injection rate that matched the processor clock frequency and sufficient bandwidth to sustain a reference from every processor on every clock cycle. Unfortunately, while an excellent match to extremely irregular applications, the MTA was not cost-effective on applications that could exploit locality and had very poor single-thread performance, making it a commercial failure. The Cray XMT approximates the Tera MTA-2, substantially reducing its cost but not overcoming its narrow range of applicability.

@@

Our preliminary work on Grappa shows that it runs several graph-crunching applications (classic examples of irregular behavior) efficiently on a commodity cluster.  For unbalanced tree search, Grappa is over 3X faster and shows greatly improved scalability compared to the Cray XMT; conversely, for breadth first search Grappa is 2.5X slower -- fast compared to implementations without Grappa, but obviously there is room for improvement.  Most importantly, however, for significantly less real world cost, users can \emph{add} significantly more processors to a commodity cluster than an XMT machine and use Grappa to achieve scalable performance.

@@

\begin{itemize}

\item \textbf{Improved network communication:} Our preliminary work has also revealed limitations of the current Grappa runtime system with regards to network scaling.  This proposal seeks funding to enhance the network layer and explore the hardware/software interactions of remote direct memory access (RDMA).

\item \textbf{Support for \ssd\ only memory systems:} The size of data being processed by large systems today (hundreds of terabytes) dwarfs what can fit into a single system. Multiple system nodes are required to maintain and access all of this data efficiently.  However, even with hundreds of channels to persistent storage, loading this data into DRAM can prove a bottleneck.  The Grappa runtime is designed to make applications tolerate extremely high memory latency, thus an idea we wish to explore is using solid state disks as main memory.

\item \textbf{Language and compiler support:} Presently, applications are hand-coded to the Grappa interface.  A significant research thrust of the proposed work is to ease this programming burden by language and compilation support.

\item \textbf{Comprehensive benchmark suite:} Presently there does not exist a comprehensive benchmark suite of irregular and graph applications.  We also seeking funding through this proposal to put together just such a benchmark suite.  This suite will include benchmarks, reference implementations, and input data-sets.  In addition a website will be created \texttt{graphbench.org} to distribute the benchmark suite and publish results.

@@

\begin{figure}[t]
\begin{center}     
  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
\begin{minipage}{0.95\columnwidth} 
  \caption{\label{fig:grappa} Grappa system overview}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

@@

Delegate operations are always executed at the home core of their address, and while arbitrary memory operations can be delegated, we restrict the use of delegate operations in three ways to make them more useful for synchronization. First, we limit each task to one outstanding delegate operation to avoid the possibility of reordering in the network. Second, we limit delegate operations to operate on objects in the 2D address space or objects that fit in a single block of the linear address space so they can be satisfied with a single network request. Finally, no context switches are allowed while the data is being modified. Given these restrictions, we can ensure that delegate operations for the same address from multiple requesters are always serialized through a single core in the system, providing atomic semantics without using atomic operations.

@@

\comment{I really don't know if this is true or if we will be able to to know that is true any time soon -Mark

The final challenge is that atomic operations via RDMA do not interact atomically with processors in a system.  Semantically it is not entirely clear what happens if a processor issues an atomic operation on a memory address at the same time the network card is attempting to service a remote atomic access.
}

@@

Grappa applications are likely to use large amounts of data, 100s of terabytes to petabytes. Keeping all these data in DRAM only is unlikely to be cost effective. We will explore mechanisms that allow data to reside in solid-state storage {\em only}.

Grappa is very effective in hiding latencies of network communication.  If an access is to a chunk of the heap stored in a node's local memory, Grappa issues a prefetch for the address and context switches to other work while the prefetch is executing~\cite{Nelson:hotpar2011}. If the access is to memory stored on another node, Grappa issues a message to that node to request the data and context switches to other work until the reply arrives.  In fact, our measurements and estimates show that we might be able to tolerate even greater latency -- on the order of milliseconds per request.  This provides the tantalizing possibility that we will be able to keep data in \ssd\ drives only and service data access requests directly from them, without leaving copies in DRAM.  The natural point of integration for non-volatile memories in Grappa is to store the global distributed shared heap in non-volatile memory. All accesses to this shared heap are done through an \api, so this switch would be transparent to the programmer.

This is not straightforward, however, for three reasons: (1) \ssds\ assume spatial locality. In our benchmarks, we find the average request size to be less than 64 bytes. Today's \ssds\ are optimized for reads in the multi-kilobyte range. (2) \ssds\ have a low request rate. The fastest~\cite{fusionio} can perform on the order of 1 million random reads per second, and many are limited to less than 100K/s. Even the underlying flash chips are slow, on the order of 40K/s~\cite{micronFlash}. In contrast, our measurements suggest cores may be idle in Grappa if each node cannot perform 100 million random reads per second of the global heap. And (3) \ssds\ have high overhead. Mass market \ssds\ are usually treated as normal disks: accesses must occur through the kernel, incurring a high context-switch overhead. Grappa depends on user-mode access to the network layer to avoid this overhead; a similar approach has been applied to \ssds\ in recent research~\cite{caulfield:2012}, however.  We see several ways to address these challenges:

\paragraph{Scaling up the Flash array:} Existing Flash-based \ssds\ are designed for large block requests. An \ssd\ for Grappa could be designed to support high concurrency, small requests. Where a traditional \ssd\ might put multiple Flash chips on the same bus, or use Flash chips in parallel to build a wider, high-bandwidth bus, a Grappa \ssd\ would try to maximize message rate by putting each chip on its own bus. Serial Flash might be used to minimize pin count requirements, at the cost of some additional latency. This design could scale to our necessary request rates, but only at an enormously high chip count: to support 100 million reads per second, an array of 25$\mu$s flash would require 2500 chips.

\paragraph{Scaling down the processors:} Alternatively, rather than scaling up the flash array, processors could be scaled down to match the flash request rate. A low-power processor with few cores might generate only 1 million references per second. This could be satisfied with only $25\times 25\mu$s Flash chips. There are two problems with this design for our purposes, however. First, while each node requires fewer chips, the overall cluster performance goal, and thus chip count requirement, is the same. More system nodes are thus required to achieve the same level of performance.  Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.

\paragraph{Buffering requests to increase locality:} Grappa trades latency for throughput by aggregating network communication. The same idea could be applied to increase locality and amortize the cost of Flash accesses. Requests to nearby flash locations would be bucketed by flash page address for a limited time. When a sufficient number of requests have arrived for a page or when the requests time out, a single page request would be issued to the Flash layer; its result would be used to satisfy the individual requests.  The challenge with this approach is as storage size grows, the effectiveness of aggregation correspondingly decreases.  On the other hand, aggregation occurs for all global memory requests generated by any node in the system, so as aggregate system performance increases, aggregation performance actually \emph{increases} (up to the point that the network is saturated).

\paragraph{Rearchitecting flash:} Clearly the best solution here lays in re-architecting the flash chip itself.  We will explore new flash microarchitetures that optimize for message rate as well as bandwidth.  Such chips have value far beyond Grappa.  Research into persistent key/value stores, and really any persistent high-throughput data-structure will benefit from a redesigned flash chip that can sustain dramatically increased message rates.

@@

Language and compiler support are instrumental in making Grappa usable to a wide audience of software developers. We envision language and compiler support can tackle a wide variety of challenges:

\paragraph{Relaxed consistency:} Presently Grappa uses a consistency model that is sequentially consistent for data-race free programs.  Underlying hooks are available to build even more relaxed consistency models, although we don't expect programmers themselves will want to use them.  A compiler, however, can see a performance boost by analyzing code and exploiting alias knowledge to program to this underlying Grappa \api, potentially gaining a performance boost by exposing more memory concurrency.

\paragraph{Language support for synchronization operations: } At the moment, runtime library support is provided in Grappa for basic synchronization primitives (lock, barrier, etc).  Building synchronization directly into the language has the advantage that the compiler can understand and optimize it.  Several optimizations are possible in the Grappa framework, including exploiting the scheduler and work stealer to elide certain lock operations (scheduling is co-operative on a per-CPU basis), converting synchronization primitives into remote delegate operations, and exploiting the check-in / check-out nature of global memory in Grappa to coalesce multiple synchronization operations together.

\paragraph{Delayed writes: } At the moment when we we write applications for Grappa we use a technique that we call feed-forward delegates.  Essentially what this technique does is if a loop is reading and writing disjoint sets of data, it is possible and we have found efficient, to dispatch writes asynchronously and ensure those writes are completed only by the end of the loop.  This increases performance by further overlapping computation with memory access.  A compiler with alias analysis can do this transformation automatically for the programmer.

\paragraph{Read and write combining: } When multiple remote memory operations are to be dispatched to the same node from the same thread, it is efficient to combine these together into a single request.  We will develop a compiler pass to combine operations when possible.

\paragraph{Automatic delegate synthesis: } Delegates are a technique in Grappa where one thread can request a small computation to be carried out on a remote node.  For instance, an increment on a memory location is dispatched to the system node that manages that memory operation and carried out there.  Currently the decision of what is a delegate is left to the programmer.  But this is an area rich for compiler optimization.  We will explore techniques to automatically synthesize delegates from code, potentially leading to more complex delegates and more efficient code.

\paragraph{Compile for a continuations model: } Presently, whenever a task is active it is allocated a complete stack.  Since several thousand tasks are active per processor at any one time, these stacks occupy a large footprint in memory.  We wish to explore compiling tasks to a continuation model, whereby when a task dispatches a remote memory operation and is to be contextd out, another task can utilize that stack.  Combined with the delegate synthesis ideas described above, we will also explore runtime and compiler suppor to convert remote memory operations into thread migrations; instead of moving the data to the system node that has the task, we will move the task to the data in cases where that would be more efficient.

\paragraph{Task-weaving: } Much of the time the same task code is executing for hundreds of thousands of tasks in the Grappa runtime system.  For example, each task may be processing a loop body for a parallel for loop.  This is not always the most efficient approach to execution, and we already manually do recursive decomposition of parallel loops with a cut-off to group a collection of loop bodies together into a single task.  The task then executes each of these loop bodies sequentially.  This is not the most efficient approach, however.  An even more efficient technique is for the compiler to unroll and schedule all loop bodies simultaneously within the same task, thereby ``weaving'' these independent tasks together more efficiently to exploit computation / communication overlap.

Presently we are starting work on this language and compiler support.  We are basing our effort on the LLVM~\cite{llvm} infrastructure.

\comment{ three main aspects: (1) easing the interface to the Grappa memory model which is sequentially consistent for data-race free programs; (2) automatic insertion of calls to the runtime to minimize the complexity of accessing globally shared memory; and (3) performing code optimizations specific for Grappa as well as traditional optimizations in ways that will not break program semantics.

We will develop type qualifiers to let programmers declare shared data, e.g., \texttt{shared int}. This will enable the compiler to automatically insert runtime calls for accesses to global heap data as well synchronized operations (e.g. atomic fetch-and-add) to global variables. This will also enable new analyses and optimizations, for example, the compiler could perform spatial and temporal locality analysis to improve request aggregation when it can prove that several accesses go to the same destination.

}

\comment{Luis, don't rant.    -mark

Specifying the memory model and access semantics of the Grappa runtime is very important for several reasons. The programmer needs to know the behavior of accesses to the global heap. The compiler also must rely on the precise access semantics so that optimizations can be performed without breaking program semantics. We opted for a sequential consistency for data-race-free programs, similarly to the C++0x memory model (something programmers and system software designers are already used to).
}

@@

In order to facilitate the development and evaluation of Grappa, we will develop a comprehensive benchmark suite for graph analytics applications. This will not only be useful for us, but the whole community at large, since there is a lack of benchmarks in this application space.

This benchmark suite will consist of application kernels, reference implementations, data-sets, data-set generators, documentation, specifications for generating standardized performance results, and a hosting service (http://graphbench.org) to provide public access to the benchmark and datasets as well as aggregate and publish results by all users of the benchmark suite.

This benchmark suite will be a tremendous value to the research community. Not only will it enable researchers to evaluate ideas on graph processing system techniques, but it will also enable fair comparisons of ideas in a depth not currently available from other graph related benchmark suites. The benchmark will be released in a completely open fashion, so anyone will be able to use it (and contribute to it).  All code will be distributed into the public domain (BSD style licensing) so companies can use it. The goal is to get something out relatively early in the hope of (i) getting others to cluster around this work rather than develop alternatives, and (ii) help establish consistent metrics for performance results.

We will build a collection of key computational kernels from graph and irregular applications. The benchmark suite will consist of both documentation on the actual kernel (ideally a research paper or book reference sourced from the community), and a reference implementation. The benchmark suite will also include sample data collected from existing sources, and synthetic workload generators. The collected data provides a grounding for performance analysis on real-world problems. The workload generators provide cleaner, easier to reason about randomly generated data-sets at a programmable scale.

\section{Broader impacts}

\paragraph{Increasing involvement of underrepresented groups:} Computer science as a whole has a significant problem attracting women and minorities to the field.  Computer architecture has an even more extreme problem with diversity (in the PI's view).  This is a problem the PIs have taken quite seriously since entering the field, and we are quite proud of our PhD students, both female who have gone on to faculty positions in top academic institutions, and male who appreciate the gender and ethnic disparity and who work hard at their own institutions to correct it.  We continue to recruit a diverse graduate student mix and we continue to include undergraduate students in our research.  At the moment we are mentoring two women undergraduates involved in the Grappa project, and if this proposal is funded will apply for REU funds to continue their involvement.

\paragraph{Benchmark suite:} As previously described, we are seeking funding support to build and distribute a benchmark suite for irregular and graph applications.  This benchmark suite has the potential to have a broad impact on academia and industry.  We will distribute this benchmark through \texttt{graphbench.org} and will also collect and publish results from competing research groups.

\paragraph{Technology transfer:} Bill Dally during his 2010 Eckert-Mauchly award speech said (in effect), technology transfers happen when people transfer.  Having just come back from a three year stint founding a startup company (where we transferred the Deterministic Multiprocessing technology from UW), we've seen this first hand.  The students and faculty involved in the Grappa project will spin off a company with their work.  The exact time frame is not yet known, but everyone involved in the project is excited to see the technology underpin a startup company and make this technology available to the broader commercial world.

\paragraph{Curriculum development activities:} Concurrent with the execution of the research activities described in this grant we will seek to teach an advanced graduate class on advanced concurrency concepts.  We taught a similar class over six years ago and it is time to revive, update and teach it to the next generation of computer architecture graduate students.  \comment{Add more here.  This has no passion. -Mark}

\section{Results from Prior NSF Support}

\paragraph{Ceze:} Prof. Ceze's research is on computer architecture, programming languages and operating systems. His primary focus has been on improving programmability, reliability and energy efficiency of multiprocessor computer systems. His research is funded primarily by NSF, PNNL and industry. His CAREER award {\em Deterministic Shared Memory Multiprocessing: Vision, Architecture and Impact on Programmability} (CCF-0846004, 2008-13, \$559K) demonstrated that it is possible to provide deterministic execution of arbitrary multithreaded programs without sacrificing performance significantly. The results include innovation in architecture, compiler, and operating-systems. Two other NSF grants, {\em Code-Centric Approach to Specifying, Checking, and Discovering Shared-Memory Communication} (CCF-1064497, 2011-2014, \$901K) and {\em Precise Concurrency Exceptions: Architecture Support, Semantics and System Implications} (CCF-1016495, 2010-2013, \$500K), are leading to significant results in dealing with concurrency errors and concurrent programming language semantics. His work on large-scale graph processing is being funded by the Pacific Northwest Laboratory. His work on energy-aware programming models and approximate computing has been primarily funded by a seed NSF grant {\em Disciplined Approximate Programming for Energy-Efficient Computing} (CCF-1216611, 2012-2015, \$300K) and industry, including Microsoft, Google and Qualcomm.

\paragraph{Kahan:} Dr. Kahan's research has focused on parallel computation runtimes, for irregular computation and biological applications.  He is presently an employee of the Department of Energy.  He has no past or current NSF support.

\paragraph{Oskin:} Prof. Oskin's prior research has focused on architectures and systems for scalable computing.  With support from NSF he has carried out a variety of research activities, leading to over 50 scholarly publications, three PhD students graduated, four masters students graduated, and a company spin-off.  The past NSF-supported work (WaveScalar) heavily influences the ideas described here. In some sense, the work here is a culmination in thought processes about why WaveScalar did not achieve as much ILP as earlier limit studies in ILP suggested were possible.  NSF support for the PI's work, {\em ITR: WaveScalar, A New Approach to Scalable System Design} (CCF-0325635, 2003-07, \$1.3M), {\em CAREER: Soft-Instruction Set Computing} (CCF-0133188, 2002-05, \$443K), helped to develop WaveScalar.  He has also spent over a decade exploring architectural implications of quantum computing technologies, with support from NSF {\em NER: Computer Aided Design of Silicon-based Quantum Computers} (CCF-0210373, 2002-04, \$90K), {\em EMT: Microarchitectures for Quantum Computers} (CCF-0523359, 2006-10, \$275K), {\em EMT: Self-Correcting Fault-Tolerant Quantum Computers} (CCF-0523359, 2005-08, \$406K) DARPA {\em QuIST: Architectures and Applications for Scalable Quantum Information Systems}.  He is currently exploring runtime systems for irregular parallel applications, which is the subject of this proposal and is partially funded by the Department of Energy.

\section{Summary}

We are awesome.  Give us money.
